# Results

```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(ggridges)
library(forcats)
library(vcd)
library(grid)
library(ggalluvial)
library(RColorBrewer)
```


```{r}
car_crash <- read.csv("crash2.csv")
car_crash[car_crash == ""] <- NA
car_crash[car_crash == "N/A"] <- NA
head(car_crash)
```



## Understanding Collision and Route Types

To start, we explore the overall distribution of collision types to understand the major categories of traffic incidents. This foundational view helps us identify which types of collisions are most prevalent.

```{r}
# Bar Plot (Frequency of Collision Types)

# Custom color scheme
colors_for_collisions <- c(
  "same direction collisions" = "#E41A1C",   # Red
  "single vehicle collisions" = "#377EB8",   # Blue
  "head-on collisions" = "#4DAF4A",          # Green
  "sideswipe collisions" = "#984EA3",        # Purple
  "unknown" = "#FF7F00",                     # Orange
  "front/rear collisions" = "#FFFF33",       # Yellow
  "turning collisions" = "#A65628",          # Brown
  "opposite direction collisions" = "#F781BF" # Pink
)

# Group and count the Collision.Type values
collision_counts <- car_crash |> 
  group_by(Collision.Type) |> 
  summarise(Count = n())

# Create the bar graph with the custom color scheme
ggplot(collision_counts, aes(x = reorder(Collision.Type, -Count), y = Count, fill = Collision.Type)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  scale_fill_manual(values = colors_for_collisions) +
  labs(
    title = "Count of Collision Types",
    x = "Collision Type",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5, size = 16)
  )

```
This bar chart displays the counts of different collision types in the dataset. The most frequent collision type is same direction collisions, followed by single vehicle collisions. Less common types include turning collisions, front/rear collisions and opposite direction collisions. A notable portion of collisions is categorized as unknown, which may require further investigation or clarification. This distribution highlights that many crashes involve vehicles traveling in the same direction, emphasizing potential issues such as tailgating or lane changes.

Next, we examine the types of routes where these collisions occur most frequently. By linking collision types to specific route categories, we gain insights into the environments contributing to these incidents.


```{r}
# Bar Plot (Frequency of Route Types)
crash2_data <- read.csv("crash2.csv")
crash2_summary <- crash2_data %>%
  filter(Light %in% c("daylight", "dark lights on", "dark no lights", "dusk", "dawn")) %>% # Select relevant light conditions
  count(Light, ACRS.Report.Type) # Count occurrences by light and report type


# Define custom color scheme
colors_for_route_type <- c(
  "unknown" = "#FF7F00",                 # Orange
  "maryland (state)" = "#E377C2",       # Yellow
  "county" = "#66C2A5",                  # Teal
  "municipality" = "#8DA0CB",            # Mauve
  "interstate (state)" = "#2CA02C",      # Dark Green
  "us (state)" = "#E6550D",              # Burnt Orange
  "other public roadway" = "#A63603",    # Brick Red
  "government" = "#08519C",              # Navy Blue
  "ramp" = "#17BECF",                    # Dark Cyan
  "service road" = "#8C564B",            # Maroon
  "municipality route" = "#D62728",      # Magenta
  "spur" = "#6A51A3",                    # Slate Blue
  "local route" = "#228B22",             # Forest Green
  "bicycle route" = "#663399",           # Dark Purple
  "private route" = "#FFFF33",           # Bright Pink
  "crossover" = "#BDB76B"                # Dark Yellow
)

# Count occurrences of each Route.Type in crash2_data
route_counts <- crash2_data %>%
  group_by(Route.Type) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count))

# Create the bar graph with custom color scheme
ggplot(route_counts, aes(x = reorder(Route.Type, -Count), y = Count, fill = Route.Type)) +
  geom_bar(stat = "identity", show.legend = FALSE, width = 0.7) +  # Narrow bars for better distinction
  geom_text(aes(label = Count), vjust = -0.5, size = 4, color = "black") +  # Add labels on top of each bar
  labs(
    title = "Count of Route Types",
    x = "Route Type",
    y = "Count"
  ) +
  scale_y_continuous(expand = c(0, 0), limits = c(0, max(route_counts$Count) * 1.1)) +  # Add space for labels
  scale_fill_manual(values = colors_for_route_type) +  # Use custom colors
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),  # Rotate x-axis labels
    plot.title = element_text(size = 16, hjust = 0.5, face = "bold"),
    axis.title = element_text(size = 12)
  )
```
This bar chart illustrates the distribution of traffic collisions across different route types. The most common route types are Maryland (state) and county roads, which together account for a majority of the reported collisions. Unknown route types also make up a significant portion, indicating missing or unclear data. Other route types, such as municipality roads, US (state) roads, and interstate highways, have lower counts but still contribute meaningfully to the overall distribution. These findings highlight that state and county roads are the primary locations for traffic incidents, emphasizing the importance of focusing safety measures and traffic management efforts on these areas.


To combine these insights, we explore how specific collision types vary across different route types. This visualization reveals patterns unique to certain routes, such as the dominance of 'single vehicle collisions' on rural roads or 'front/rear collisions' on urban streets.

```{r fig.width=12, fig.height=8, fig.align="center"}
# Bar Plot (Collision Types faceted by Route Type)
# Define updated custom color scheme
colors_for_route_type <- c(
  "unknown" = "#FF7F00",                 # Orange
  "maryland (state)" = "#E377C2",        # Bright Pink
  "county" = "#66C2A5",                  # Teal
  "municipality" = "#8DA0CB",            # Mauve
  "interstate (state)" = "#2CA02C",      # Dark Green
  "us (state)" = "#E6550D",              # Burnt Orange
  "other public roadway" = "#A63603",    # Brick Red
  "government" = "#08519C",              # Navy Blue
  "ramp" = "#17BECF",                    # Dark Cyan
  "service road" = "#8C564B",            # Maroon
  "municipality route" = "#D62728",      # Magenta
  "spur" = "#6A51A3",                    # Slate Blue
  "local route" = "#228B22",             # Forest Green
  "bicycle route" = "#663399",           # Dark Purple
  "private route" = "#FFFF33",           # Yellow
  "crossover" = "#BDB76B"                # Dark Yellow
)

# Prepare data, sort Route.Type by Collision.Count in descending order within each Collision.Type
collision_summary <- car_crash |> 
  filter(!is.na(Route.Type), !is.na(Collision.Type)) |> 
  group_by(Route.Type, Collision.Type) |> 
  summarize(Collision.Count = n(), .groups = 'drop') |> 
  arrange(Collision.Type, desc(Collision.Count))

# Reorder Route.Type within each Collision.Type
collision_summary$Route.Type <- factor(collision_summary$Route.Type, 
                                       levels = unique(collision_summary$Route.Type))

# Create the plot
ggplot(collision_summary, aes(x = Route.Type, y = Collision.Count, fill = Route.Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Collision Types by Route Type",
    x = "Route Type",
    y = "Number of Collisions",
    fill = "Route Type"
  ) +
  scale_fill_manual(values = colors_for_route_type) +  # Use the custom color scheme
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold")
  ) +
  facet_wrap(~ Collision.Type, scales = "free_y", ncol = 4)

```

The Maryland (state) category and county category exhibit the highest number of collisions across nearly all types, including single vehicle collisions, head-on collisions, and same direction collisions, suggesting that local roads are particularly prone to accidents. This pattern may result from higher traffic volumes, frequent stops, or more complex navigation on local roads. County roads also account for a significant number of collisions, especially single vehicle collisions and same direction collisions, though the majority of collisions still occur on local routes. Notably, opposite direction collisions are predominantly observed on county routes, which are often characterized by two-lane undivided structures and higher speed limits, increasing the risk of head-on crashes. This emphasizes the need for targeted safety measures such as centerline rumble strips, enhanced signage, and improved road design to reduce risks on county roads.
Collisions under the "unknown" category are disproportionately high on unknown route types, suggesting potential issues in data collection or reporting that could hinder the accurate analysis of trends. In contrast, bicycle routes, spurs, service roads, and private roads report the lowest collision counts, likely reflecting lower usage or dedicated infrastructure for these road types. These findings highlight the importance of prioritizing safety interventions and infrastructure improvements, particularly on local and county roads where collisions are most prevalent.


## The Role of Traffic Controls

Traffic control systems play a crucial role in managing safety on the road. Here, we analyze the distribution of traffic controls across incidents to understand their presence or absence in various collision scenarios.
```{r}
# Bar Plot (Frequency of Traffic Control Types)

car_crash <- read.csv("crash2.csv")

# Count occurrences of each Traffic.Control
traffic_control_counts <- car_crash |>
  group_by(Traffic.Control) |>
  summarise(Count = n()) |>
  arrange(desc(Count))

colors_for_traffic_control <- c(
  "no controls" = "#A6CEE3",   
  "traffic signal" = "#FB9A99",   
  "unknown" = "#807DBA",                  
  "stop sign" = "#B3DE69",                
  "flashing traffic signal" = "#BCBD22",
  "yield sign" = "#9E9AC8",              
  "person" = "#FDAE6B",        
  "warning sign" = "#31A354",      
  "other signal" = "#9C9EDE",    
  "pedestrian crossing" = "#F768A1",   
  "railroad crossing device" = "#F7FB6B",
  "other pavement marking" = "#7FCDBB", 
  "school zone sign" = "#B2DF9A", 
  "bicycle crossing sign" = "#969696"
)

# Create the bar graph
ggplot(traffic_control_counts, aes(x = reorder(Traffic.Control, -Count), y = Count, fill = Traffic.Control)) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  scale_fill_manual(values = colors_for_traffic_control) + 
  labs(
    title = "Count of Traffic Control Types",
    x = "Traffic Control Type",
    y = "Count"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),  # Rotate x-axis labels
    plot.title = element_text(size = 16, hjust = 0.5),
    axis.title = element_text(size = 12)
  ) 

```
The bar chart reveals that most crashes occurred in areas with no traffic controls, highlighting a significant safety concern. Traffic signals are the second common incidents, indicating that while they regulate traffic, they remain accident-prone due to possible human or mechanical errors. Stop signs also contributed to crashes but at a much lower frequency. Other measures, such as flashing signals and yield signs, showed minimal crash counts, suggesting either limited usage or greater effectiveness. This analysis emphasizes the need for enhanced safety measures, especially in uncontrolled areas, to reduce accidents and improve road safety for all users.


Before analyzing the influence of traffic controls on crash outcomes, we take a closer look at the overall distribution of crash types. This provides essential context for understanding how different crash severities are impacted by traffic management.

```{r}
# Bar Plot (Frequency of Car Crash Types)

car_crash_counts <- car_crash |>
  group_by(ACRS.Report.Type) |>
  summarise(Count = n()) |>
  arrange(desc(Count))

colors_for_crash_type <- c(
  "fatal crash" = "#CC7A8B",
  "injury crash" = "#7FBF7F", 
  "property damage crash" = "#89CFF0"
)
ggplot(car_crash_counts, aes(x = reorder(ACRS.Report.Type, -Count), y = Count, fill = ACRS.Report.Type)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = colors_for_crash_type) +
  labs(
    title = "Frequency of Crash Types",
    x = "Crash Type",
    y = "Count"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

This bar plot visualizes the distribution of crash types in the dataset, categorized into three main types: property damage crashes, injury crashes, and fatal crashes. The plot reveals that property damage crashes are the most frequent, significantly outnumbering the other two types. Injury crashes are the second most common, while fatal crashes occur relatively infrequently. The stark difference in counts highlights the prevalence of less severe accidents compared to life-threatening or fatal incidents. This foundational analysis provides critical context for further exploration into the factors that influence crash severity.


To evaluate the effectiveness of traffic controls, we explore their relationship with crash severity. This mosaic plot highlights how the presence of stop signs or traffic signals correlates with fewer severe crashes.


```{r}
# Mosaic Plot (Crash Type & Traffic Control)
# Define the custom color scheme for Traffic Control
colors_for_traffic_control <- c(
  "no controls" = "#A6CEE3",   
  "traffic signal" = "#FB9A99",   
  "stop sign" = "#B3DE69"
)

# Add a new column for selected traffic controls
car_crash$Traffic.Control.selected <- ifelse(
  car_crash$Traffic.Control %in% names(colors_for_traffic_control),
  car_crash$Traffic.Control,
  NA  # Set to NA if it doesn't match the selected categories
)

# Remove rows where Traffic.Control.selected is NA
car_crash <- car_crash[!is.na(car_crash$Traffic.Control.selected), ]

# Generate a mosaic plot with Traffic Control on y-axis and Report Type on x-axis
mosaic(
  ~ Traffic.Control.selected + ACRS.Report.Type,  # Formula for mosaic plot
  data = car_crash,
  direction = "v",  # Vertical direction for Traffic Control as y-axis
  gp = gpar(
    fill = colors_for_traffic_control[levels(as.factor(car_crash$Traffic.Control.selected))]
  ),
  main = "Mosaic Plot: Traffic Control vs Crash Type",
  xlab = "Crash Type",
  ylab = "Traffic Control",
  labeling = labeling_values,  # Add frequencies to each part of the plot
  labeling_args = list(
    rot_labels = c(bottom = 45),  # Rotate x-axis labels to slant
    gp_labels = list(bottom = gpar(fontsize = 8, fontface = "italic")),  # Customize font for x-axis labels
    gp_text = gpar(fontsize = 10, col = "black")  # Customize frequency label font size and color
  )
)

```

The mosaic plot shows how different types of traffic controls, such as "no controls," "stop signs," and "traffic signals", relate to the severity of crashes. Notably, property damage crashes dominate across all traffic control types. Crashes at intersections with "no controls" are the most frequent overall, especially for injury and property damage incidents, emphasizing the risks of unregulated roadways. On the other hand, crashes at "traffic signals" are less likely to result in fatalities but still significant for property damage. The low number of crashes at "stop signs" suggests their relative effectiveness in regulating traffic, though their small sample size may warrant further study. This analysis highlights the critical role of traffic control measures in reducing severe crash outcomes.



To delve deeper, we use an alluvial plot to trace the flow from traffic controls to collision types and crash outcomes. This comprehensive view uncovers trends, such as the role of 'no controls' in severe crashes.


```{r}
# Alluvial (Traffic Control & Collision Type & Crash Type)

filtered_data <- crash2_data[, c("Traffic.Control", "Collision.Type", "ACRS.Report.Type")]
filtered_data <- na.omit(filtered_data)

# Convert columns to factors
filtered_data$Traffic.Control <- factor(filtered_data$Traffic.Control)
filtered_data$Collision.Type <- factor(filtered_data$Collision.Type)
filtered_data$ACRS.Report.Type <- factor(filtered_data$ACRS.Report.Type)
filtered_data <- filtered_data %>%
  filter(
    Traffic.Control %in% c("no controls", "traffic signal", "stop sign"),
    Collision.Type %in% c("same direction collisions", "sideswipe collisions", "single vehicle collisions", "head-on collisions")
  )

# Define the custom color scheme for Traffic Control
colors_for_traffic_control <- c(
  "no controls" = "#A6CEE3",   
  "traffic signal" = "#FB9A99",   
  "unknown" = "#807DBA",                  
  "stop sign" = "#B3DE69",                
  "flashing traffic signal" = "#BCBD22",
  "yield sign" = "#9E9AC8",              
  "person" = "#FDAE6B",        
  "warning sign" = "#31A354",      
  "other signal" = "#9C9EDE",    
  "pedestrian crossing" = "#F768A1",   
  "railroad crossing device" = "#F7FB6B",
  "other pavement marking" = "#7FCDBB", 
  "school zone sign" = "#B2DF9A", 
  "bicycle crossing sign" = "#969696"
)

# Summarize the data
alluvium_data <- filtered_data %>%
  group_by(Traffic.Control, Collision.Type, ACRS.Report.Type) %>%
  summarize(Freq = n(), .groups = 'drop')

# Create the alluvium plot with the custom Traffic Control color scheme
ggplot(alluvium_data, aes(axis1 = Traffic.Control, axis2 = Collision.Type, axis3 = ACRS.Report.Type, y = Freq)) +
  geom_alluvium(aes(fill = Traffic.Control)) +  # Fill based on Traffic.Control
  geom_stratum() +
  geom_text(stat = "stratum", aes(label = after_stat(stratum))) +
  scale_x_discrete(limits = c("Traffic.Control", "Collision.Type", "ACRS.Report.Type"), expand = c(0.1, 0.1)) +
  scale_fill_manual(values = colors_for_traffic_control) +  # Apply custom colors
  labs(
    title = "Relationship between Traffic Control, Collision Type, and Report Type",
    x = "Categories",
    y = "Frequency",
    fill = "Traffic Control"
  ) +
  theme_minimal()
```
This alluvial diagram highlights the relationship between traffic control types, collision types, and crash severities. Areas with "no controls" account for the majority of single vehicle and sideswipe collisions, leading predominantly to property damage crashes and a smaller proportion of injury crashes. This pattern underscores the risks posed by unregulated roadways, especially for drivers prone to lane changes or loss of control. In contrast, "traffic signals" are primarily linked to same direction collisions, contributing significantly to property damage and injury crashes, indicating possible driver errors or violations at intersections. "Stop signs" are less frequent and associated mainly with sideswipe and same direction collisions. To mitigate these risks, implementing traffic controls, such as rumble strips or advanced warning systems, in high-risk areas could help prevent single vehicle incidents, while improving signal visibility and timing may reduce same direction collisions.


## Influence of Speed and Light Conditions

Speed limits are another critical factor influencing collision outcomes. This box plot illustrates the range of speed limits associated with different collision types, revealing that higher speeds are often linked to severe incidents.
```{r}
ggplot(crash2_data, aes(x = Collision.Type, y = Speed.Limit)) +
  geom_boxplot(outlier.color = "red", fill = "lightblue") +
  labs(
    title = "Variation of Speed Limits Across Collision Types",
    x = "Collision Type",
    y = "Speed Limit"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1) # Rotate x-axis labels for readability
  )
```
The boxplot suggests that while speed limits across collision types share some similarities, there are notable differences in the median and variability, particularly for opposite direction collisions and unknown collisions. Opposite direction collisions show a lower median speed limit, suggesting they often occur on roads with relatively lower speed limits. Other groups shows a similar median, showing that other types of collisions are most likely to occur on roads with speed limit between 30-40. 


Building on the previous analysis, this histogram focuses on crash types and their speed limit distributions. It highlights whether lower speed limits effectively reduce the severity of crashes.
```{r}
# Ridgeline Plot (Speed Limit Faceted by Crash Types)

top_report_types <- car_crash |>
  count(ACRS.Report.Type) |>
  top_n(3, wt = n) |>
  pull(ACRS.Report.Type)

filtered_data <- car_crash |>
  filter(ACRS.Report.Type %in% top_report_types)

# Ridge plot with custom colors
ggplot(filtered_data, aes(x = Speed.Limit, y = fct_reorder(ACRS.Report.Type, Speed.Limit), fill = ACRS.Report.Type)) +
  geom_density_ridges(alpha = 0.7, scale = 1.5) +
  scale_fill_manual(values = c("#CC7A8B", "#7FBF7F", "#89CFF0")) + # Custom pastel colors
  labs(
    title = "Speed Limit Distribution by ACRS Report Types",
    x = "Speed Limit (mph)",
    y = "ACRS Report Type",
    fill = "ACRS Report Type"
  ) +
  theme_minimal() +
  theme(
    legend.position = "top",
    axis.text.y = element_text(size = 10, angle = 0),
    plot.title = element_text(size = 14, hjust = 0.5)
  )

```
This ridge plot shows how speed limits are distributed across different crash severity levels: fatal crashes, injury crashes, and property damage crashes.
Fatal crashes tend to occur at higher speed limits, with a notable peak around 40 mph. Injury crashes show a broader distribution of speed limits, with a peak at approximately 35 mph. Property damage-only crashes are concentrated at moderate speed limits, particularly around 30-35 mph. These types of crashes are less severe and are often associated with lower-speed urban environments or rear-end collisions.
Higher speeds are strongly associated with more severe crashes (injuries and fatalities), while lower-speed environments tend to result in property damage-only incidents. This analysis highlights the importance of speed regulation and traffic control measures in reducing crash severity and preventing fatalities.
The ridgeline plot effectively compares the probability distributions of different crash types across speed limits although showing some multi-modality. It highlights the speed ranges with higher likelihoods for specific crash types.



Environmental factors, such as lighting, also play a role in crash severity. This stacked bar chart shows the distribution of crash types under different light conditions, emphasizing the higher frequency of incidents in daylight and 'dark-lighted' scenarios.




